---
title: "Darwin Core mapping"
subtitle: "For: monitoring of invasive alien crayfishes in the Flemish part of the LIFE RIPARIAS areas"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

Install required packages (only if the packages have not been installed before):

```{r install_pkgs, warning = FALSE}
installed <- rownames(installed.packages())
required <- c("tidyverse",
              "tidylog",
              "here",
              "readxl",
              "sf",
              "DBI",
              "RSQLite",
              "digest",
              "testthat"
)
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load packages:

```{r load_pkgs, warning = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(here)           # To find files
library(readxl)         # To read Excel files
library(sf)             # To convert coordinate systems
library(DBI)            # To work with databases
library(RSQLite)        # To work with SQLite databases in R
library(digest)         # To create hashes
library(testthat)       # To perform tests
```

# Read source data

Create a data frame `occurrences` from the source data:

```{r read_data}
occurrences <- readxl::read_excel(
  path = here("data", "raw", "rivierkreeften_ANB september 2022.xlsx")) %>%
  mutate(datum = as.character(.data$datum),
         x = as.character(x))
```

Preview data:

```{r preview_raw_data}
occurrences %>% head(n = 5)
```

# Process source data

Define vector with the IAS crayfishes of interest:

```{r IAS_crayfishes}
ias_crayfishes <- c(
  "Pontastacus leptodactylus", 
  "Pacifastacus leniusculus", 
  "Faxonius limosus",
  "Faxonius rusticus",
  "Procambarus clarkii",
  "Procambarus acutus",
  "Procambarus fallax"
)
```

## Tidy data

Pivot data to convert the columns `Pontastacus leptodactylus`, `Pacifastacus leniusculus`, `Faxonius limosus`, `Faxonius rusticus`, `Procambarus clarkii`, `Procambarus acutus` and `Procambarus fallax` in two columns called `species` and `n`:

```{r tidy_data}
occurrences <- 
  occurrences %>%
  tidyr::pivot_longer(
    cols = one_of(ias_crayfishes), 
    names_to = "species", 
    values_to = "n"
  )
```

Set `n` to 0 if `NA`:

```{r na_to_zero}
occurrences <- 
  occurrences %>%
  mutate(n = if_else(is.na(.data$n), 0, .data$n))
```

Example tidy transformation:

```{r example_tidy_data}
occurrences %>% 
  select(locatie, species, n) %>%
  head(20)
```

## Improve location identifiers

Some location identifiers are just numbers. Notice that they all have a non empty description (column `omschrijving`):

```{r supect_locaties}
occurrences %>%
  filter(!grepl("\\D", locatie)) %>%
  distinct(locatie, omschrijving)
```

Create column `location` composed of the first capital letter of each word in `omschrijving` and the number in `locatie`:

```{r improve_locations}
occurrences <-
  occurrences %>%
  rowwise() %>%
  mutate(
    location = if_else(
      !grepl("\\D", locatie),
      paste0(
        paste0(unlist(stringr::str_extract_all(omschrijving, 
                                              pattern = "[A-Z]")),
               collapse = ""),
        locatie, collapse = ""),
      locatie
    )) %>%
  ungroup()
```

Improved locations:

```{r show_results_improved_locations}
occurrences %>%
  filter(!grepl("\\D", locatie)) %>%
  distinct(locatie, omschrijving, location)
```

Other locations should not be changed. Example:

```{r show_results_others}
occurrences %>%
  filter(grepl("\\D", locatie)) %>% 
  distinct(locatie, omschrijving, location) %>%
  head(10)
```

## Clean geographical coordinates

Some rows contain suspicious coordinates:

```{r suspicious_coords}
occurrences %>%
  filter(!grepl("\\D", locatie)) %>%
  distinct(x, y)
```

The decimal separator (`.`), if missing, has to be set after six digits from left:

```{r x_y}
# add dot after six digits
occurrences$x_improved <- gsub("^(.{6})(.*)$",
                      "\\1\\.\\2",
                      occurrences$x)
occurrences$y_improved <- gsub("^(.{6})(.*)$",
                      "\\1\\.\\2",
                      occurrences$y)
# remove duplicate dots
occurrences$x_improved <- gsub("(\\.)\\1+", "\\1", occurrences$x_improved)
occurrences$y_improved <- gsub("(\\.)\\1+", "\\1", occurrences$y_improved)
# remove dot at the end
occurrences$x_improved <- gsub("\\.$", "", occurrences$x_improved)
occurrences$y_improved <- gsub("\\.$", "", occurrences$y_improved)
```

See changes:

```{r show_changes}
occurrences %>% distinct(x, y, x_improved, y_improved)
```

## Change coordinate system

Data are provided in Belgian Lambert72 ([EPSG 31370](https://epsg.io/31370)) coordinates. DwC accepts data in WGS84 ([EPSG 4326](https://epsg.io/4326)) only.

```{r change_crs}
occurrences <- 
  occurrences %>%
  # remove x and y as they contain errors
  select(-c(x, y)) %>%
  # create duplicate of x_improved and y_improved using lambert as suffix
  mutate(x_lambert = x_improved,
         y_lambert = y_improved) %>%
  # transform to numeric
  mutate(across(ends_with("improved"), as.numeric)) %>%
  # transform to a geospatial dataframe
  st_as_sf(crs = st_crs(31370), coords = c("x_improved", "y_improved")) %>%
  # transform corodinate reference system
  st_transform(crs = 4326)
# retrieve coordinates from geospatial dataframe
coords <- as_tibble(st_coordinates(occurrences))
# convert back to standard data.frame attaching new coordinates
occurrences <- 
  as_tibble(occurrences) %>% 
  bind_cols(coords) %>%
  select(-.data$geometry)
```

```{r include = FALSE}
# remove help data.frame, not needed anymore
remove(coords)
```

Preview:

```{r}
occurrences %>% head(5)
```

## Generate hashes for species names

We generate a hash based on the species name as saved in column `species`. This is needed to create a unique `occurrenceID` of the form `eventID:hash` where `eventID` will be later defined during the mapping. As long as the species name doesn't change, the hash and so the `occurrenceID` will be stable:

```{r generate_hashes}
vdigest <- Vectorize(digest)
# Generate hashes
occurrences <-
  occurrences %>% 
  mutate(species_name_hash = vdigest(.data$species, algo = "md5"))
```

Preview:

```{r show_hashes}
occurrences %>%
  select(starts_with("species")) %>%
  distinct()
```

# Create database

Create a SQLite database with the source data, so it can be queried with SQL in the next steps:

```{r create_db}
message("Create in memory SQLite database...")
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
# import occurrences
DBI::dbWriteTable(con, "occurrences", occurrences)
message("DONE")
```


# Darwin Core mapping

Create [Event](https://rs.gbif.org/core/dwc_event_2022-02-02.xml) core:

```{r event}
dwc_event_sql <- glue::glue_sql(
  readr::read_file(here::here("sql", "dwc_event.sql")),
  .con = con
)
dwc_event <- DBI::dbGetQuery(con, dwc_event_sql)
```

Create [Occurrence](https://rs.gbif.org/core/dwc_occurrence_2022-02-02.xml) extension:

```{r occurrence}
message("Map occurrences to DwC...")
dwc_occurrence_sql <- glue::glue_sql(
  readr::read_file(here::here("sql", "dwc_occurrence.sql")), 
  .con = con
)
dwc_occurrence <- DBI::dbGetQuery(con, dwc_occurrence_sql)
message("DONE")
```

# Save data to CSV

```{r save_csv}
write_csv(dwc_event, here::here("data", "processed", "event.csv"), na = "")
write_csv(dwc_occurrence, here::here("data", "processed", "occurrence.csv"),
          na = ""
)
```

# Test output

Load tests and run them to validate the DwC mapping:

```{r run_tests, message = TRUE}
source(here("tests", "test_dwc_event_occurrence.R"))
```

